A Large-scale Study of Polymorphism in R

### 2019-04-02 paper/main.tex
merging paper

\newcommand{\PACKAGES}{11,463\xspace}
\newcommand{\PROGRAMMERS}{?\xspace}
\newcommand{\PERCENTCRAN}{83\%\xspace}
\newcommand{\CRANTOTAL}{13,841\xspace}
\newcommand{\RLOC}{15,050,267\xspace}
\newcommand{\CLOC}{9,373,542\xspace}
\newcommand{\YEARS}{?\xspace}
\newcommand{\INDEXCOINCIDENCE}{3,499\xspace}
\newcommand{\TOTALINDEXY}{6,561\xspace}
\newcommand{\INDEXYPERC}{53\%\xspace}		% TODO Not an accurate count.
\newcommand{\DATAPKGS}{206\xspace}
\newcommand{\DATAPKGSPERC}{1.5\%\xspace}
\newcommand{\METAARGCOUNT}{7,051\xspace}

@@

	\item if we roll NULLs into every type, we can cleanly deal with these X, NULL signatures.
	As we have established that an annotation for an optional argument is useful, this will allow us to reveal other interesting patterns.
	\AT{If we feel the need to go into more details on NULL or NA, I have the old paragraphs commented out in the source.}

@@

There are some clear limitations to the L0 type system that we should address before working through any of these other issues
As we mentioned, \code{typeof} makes no distinction between vectors and scalars, as indeed there is no such distinction in R. 
But as we saw, we might benefit from such a distinction, and indeed an all-new runtime environment design for R might benefit from e.g. not needing to vectorize scalar values.
This is one of the many modifications we make, all of which are thoroughly explored in the following section.

%In terms of programmer behavior, we observe a pattern of optional arguments throughout several R packages, wherein programmers specify a default argument value which is sometimes NULL.
%Consider the snippet in Figure~\ref{fig:optnull} for an example.
%\AT{TODO put that figure here}
%The distinction between NULL and other types is not particularly useful to the programmer, so we propose to also collapse this distinction.
%Not pictured in Table~\ref{tab:toppolyFDTS} are NAs, which are ascribed the type {\tt raw\_NA} in L1; NAs in fact inhabit every type, though the default NA type is {\tt logical}.
%Just as we fold NULL into all types, we will do the same with NAs.

%
%
%
%
\subsection{L1}

@@

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 22 18:03:30 2019
\begin{table}[ht]
\label{tab:argcountsL1p}
\centering
\begin{tabular}{lrr}
  \hline
 Argument Polymorphism & Count & Percentage \\ 
  \hline
  Full Monomorphic & 781K & 85.47 \\
  Monomorphic in Type & 867K & 95.01 \\
%  Monomorphic in Class & 831K & 90.31 \\ 
%  Monomorphic in Attribute Pattern & 830K & 90.25 \\ 
  Total Seen & 914K & --- \\ 
     \hline
\end{tabular}
\caption{Account of {\it argument} polymorphism in L$1^{+}$.}
\end{table}

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 22 18:10:46 2019
\begin{table}[ht]
\label{tab:funcountsL1p}
\centering
\begin{tabular}{lrr}
  \hline
 Function Polymorphism & Count & Percentage \\ 
  \hline
  Full Monomorphic & 143K & 66.04 \\ 
  Monomorphic in Type & 184K & 85.30 \\
%  Monomorphic in Class & 163K & 75.13 \\ 
%  Monomorphic in Attribute Pattern & 161K & 74.31 \\ 
  Total Seen & 216K & --- \\ 
   \hline
\end{tabular}
\caption{Account of {\it function} polymorphism in L$1^{+}$.}
\end{table}

@@

\AT{Eyy time for L2.}

%
%
%
%
\AT{We might want to discuss some of the most polymorphic functions in the language.}
\subsection{R's Most Polymorphic Functions}

>>>>

\newcommand{\PACKAGES}{11,463\xspace}
\newcommand{\PROGRAMMERS}{?\xspace}
\newcommand{\PERCENTCRAN}{83\%\xspace}
\newcommand{\CRANTOTAL}{13,841\xspace}
\newcommand{\RLOC}{15,050,267\xspace}
\newcommand{\CLOC}{9,373,542\xspace}
\newcommand{\YEARS}{?\xspace}
\newcommand{\INDEXCOINCIDENCE}{3,499\xspace}
\newcommand{\TOTALINDEXY}{6,561\xspace}
\newcommand{\INDEXYPERC}{53\%\xspace}		% TODO Not an accurate count.
\newcommand{\DATAPKGS}{206\xspace}
\newcommand{\DATAPKGSPERC}{1.5\%\xspace}
\newcommand{\METAARGCOUNT}{7,051\xspace}
\newcommand{\NUMPOLYTYPEMONOCLASS}{1,357\xspace}

@@

	\item if we roll NULLs into every type, we can cleanly deal with these X, NULL signatures.
	As we have established that an annotation for an optional argument is useful, this will allow us to reveal other interesting patterns.

@@

There are some clear limitations to the L0 type system that we should address before working through any of these other issues
As we mentioned, \code{typeof} makes no distinction between vectors and scalars, as indeed there is no such distinction in R. 
But as we saw, we might benefit from such a distinction, and indeed an all-new runtime environment design for R might benefit from e.g. not needing to vectorize scalar values.
This is one of the many modifications we make, all of which are thoroughly explored in the following section.
%
%
%
%
\subsection{L1}

@@

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 22 18:03:30 2019
\begin{table}[ht]
\label{tab:argcountsL1p}
\centering
\begin{tabular}{lrr}
  \hline
 Argument Polymorphism & Count & Percentage \\ 
  \hline
  Full Monomorphic & 781K & 85.47 \\
  Monomorphic in Type & 869K & 95.01 \\
%  Monomorphic in Class & 831K & 90.31 \\ 
%  Monomorphic in Attribute Pattern & 830K & 90.25 \\ 
  Total Seen & 914K & --- \\ 
     \hline
\end{tabular}
\caption{Account of {\it argument} polymorphism in L$1^{+}$.}
\end{table}

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 22 18:10:46 2019
\begin{table}[ht]
\label{tab:funcountsL1p}
\centering
\begin{tabular}{lrr}
  \hline
 Function Polymorphism & Count & Percentage \\ 
  \hline
  Full Monomorphic & 143K & 66.04 \\ 
  Monomorphic in Type & 185K & 85.30 \\
%  Monomorphic in Class & 163K & 75.13 \\ 
%  Monomorphic in Attribute Pattern & 161K & 74.31 \\ 
  Total Seen & 216K & --- \\ 
   \hline
\end{tabular}
\caption{Account of {\it function} polymorphism in L$1^{+}$.}
\end{table}

@@

In L2, we aim to incorporate R's other reflective capabilities to address some of the more subtle instances of polymorphism.

We observe that the act of ascribing a class to a value in R carries the intent of the programmer.
Now, not all classes are user-defined, and R has a number of built-in or otherwise standard classes, ranging from truly primitive classes such as ``numeric'' to the more complex ``data.frame''.
Thus, before considering the insights provided by the class of values, we transform each of R's standard classes into a ``primitive'' class, which will serve to highlight user-defined classes and those particular to some packages.

\AT{I don't know the best way to approach this subsection, so I'm just gonna start looking at some stuff and see where it leads.}

First, we consider the purpose of classes in R.
The

As before, we start with the counts, see Table~\ref{tab:classcountsL2}.

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Mon Apr  1 13:02:08 2019
\begin{table}[ht]
\label{tab:classcountsL2}
\centering
\begin{tabular}{lrr}
  \hline
Type & Count & Percentage \\ 
  \hline
Full Monomorphic & 777 & 84.96 \\ 
%  Monomorphic in Type & 869.00 & 95.01 \\ 
  Monomorphic in Class & 850 & 92.94 \\ 
%  Monomorphic in Attribute Pattern & 825.00 & 90.27 \\ 
  Total Seen & 914 & --- \\ 
   \hline
\end{tabular}
\caption{Account of class-based polymorphism in L2.}
\end{table}

Top signatures in Table~\ref{tab:classpolyL2}.

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Mon Apr  1 14:48:55 2019
\begin{table}[ht]
\label{tab:classpolyL2}
\centering
\begin{tabular}{lrr}
  \hline
Class & Count & Precentage \\ 
  \hline
primitive, tbl, tbl\_df & 1796 & 2.78 \\ 
  array, primitive & 1379 & 2.14 \\ 
  data.table, primitive & 777 & 1.20 \\ 
  POSIXct, POSIXt & 716 & 1.11 \\ 
  formula, primitive & 639 & 0.99 \\ 
  xts, zoo & 588 & 0.91 \\ 
  glm, lm & 494 & 0.77 \\ 
  primitive, ts & 432 & 0.67 \\ 
  mts, primitive, ts & 384 & 0.59 \\ 
  huxtable, primitive & 318 & 0.49 \\ 
   \hline
\end{tabular}
\caption{Top 10 signatures for class-based polymorphism in L2.}
\end{table}

Given the low percentages overall, we see that there are no clear patterns here.
This data does highlight some common classes across R packages:
tbl and tbl\_df are from the dplyr package, as classes for data objects.
data.table is the primary data type in the data.table package, again for data objects.
POSIXct and POSIXt are for dates.

To see how we might use class information, we should see how often a monomorphic class signature co-occurs with a polymorphic type signature.
We count \NUMPOLYTYPEMONOCLASS argument positions which match this criteria.

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Mon Apr  1 16:50:22 2019
\begin{table}[ht]
\centering
\begin{tabular}{llrl}
  \hline
  Type Sig. & Class Sig. & Count & Percentage \\ 
  \hline
  list$<$any$>$, list$<$character$>$ & roxy\_block & 38 & 2.80 \\ 
  list$<$any$>$, list$<$list$>$ & Rd & 25 & 1.84 \\ 
  list$<$any$>$, list$<$real$>$ & TSdata & 25 & 1.84 \\ 
  list$<$any$>$, list$<$list$>$ & xml\_nodeset & 24 & 1.77 \\ 
  list$<$any$>$, list$<$list$>$, list$<$real$>$ & dendrogram & 24 & 1.77 \\ 
  list$<$any$>$, list$<$real$>$ & phylo & 23 & 1.69 \\ 
  list$<$any$>$, list$<$list$>$, list$<$real$>$, scalar/real & dendrogram & 21 & 1.55 \\ 
  list$<$any$>$, list$<$language$>$ & uneval & 18 & 1.33 \\ 
  list$<$any$>$, list$<$list$>$ & dendrogram & 18 & 1.33 \\ 
  list$<$any$>$, list$<$list$>$ & lazy\_dots & 17 & 1.25 \\ 
   \hline
\end{tabular}
\end{table}

\AT{I don't think this table is telling us much...}

Top attributes, maybe?

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Mon Apr  1 19:21:13 2019
\begin{table}[ht]
\centering
\begin{tabular}{lrl}
  \hline
attr & count & perc \\ 
  \hline
\{\}, \{\xspace names: \charv\} & 9274.00 & 10.43 \\ 
  \{\}, \{\xspace names: \chars\} & 4900.00 & 5.51 \\ 
  \{\}, \{\xspace dim: \intv\} & 4758.00 & 5.35 \\ 
  \{\xspace names: \chars\}, \{\xspace names: \charv\} & 3457.00 & 3.89 \\ 
  \{\}, \{\xspace names: \chars\}, \{\xspace names: \charv\} & 2415.00 & 2.72 \\ 
  \{\xspace dim: \intv, dimnames:list$<$any$>$ \}, \{\xspace dim:\intv\} & 2264.00 & 2.55 \\ 
  \{\}, \{\xspace srcref: \intv\} & 2037.00 & 2.29 \\ 
  \{\xspace dim: \intv, dimnames: list$<$character$>$\}, \{ dim:\intv\} & 1734.00 & 1.95 \\ 
  \{\xspace class: \chars, names: \charv, row.names:\charv\}, \{class:\chars, names:\charv, row.names:\intv\} & 1464.00 & 1.65 \\ 
  \{\xspace class: \chars, names: \charv\}, \{ class:\charv, names:\charv\} & 1277.00 & 1.44 \\ 
   \hline
\end{tabular}
\caption{Top attribute patterns. \AT{TODO fix this up}}
\end{table}\charv

A lot of the polymorphism that we see is because we haven't propagated our changes through the attribute types.
Will fix this.

%
%
%
%
\subsection{R's Most Polymorphic Functions}

\AT{We might want to discuss some of the most polymorphic functions in the language.}
